---
title: "SRC R Session 2"
author: "Yiqing Xu"
date: 
output:
  pdf_document: default
  html_document: default
---


## Resume Experiment

### Preparation
```{r}
## clean up memory, check/set working directory as necessary
rm(list = ls())
getwd()
dir()
```

### Loading and examining the data
```{r}
## load dataset
d <- read.csv("resume.csv")

## take a look at the dataset
head(d)
colnames(d)
summary(d)
unique(d$firstname) # unique first names in the data
```
\vspace{5mm}
```{r}
table(d$race)       # counts of race group for the names
```

```{r}
table(d$call)       # counts of whether or not there was a callback
```


```{r}
unique(d$firstname) # vector of unique first names in data
```

```{r}
## draw the distribution of the callback rate
hist(d$call)
```

```{r}
## how many unique first names are there in the dataset?
length(unique(d$firstname))
```

```{r}
## how many of these names are 'black' names and 'white' names, respectively?
length(unique(d$firstname[which(d$race == "white")]))
length(unique(d$firstname[which(d$race == "black")]))
```

```{r}
## how many of them are black names and white names, respectively?
length(unique(d$firstname[which(d$race == "black")]))
length(unique(d$firstname[which(d$race == "white")]))
```

```{r}
## average call rates for blacks and whites
call.wh <- mean(d$call[d$race=="white"])
call.bl <- mean(d$call[d$race=="black"])
```

```{r}
# calculate how much lower black callback rate is compared to white callback rate
(call.wh - call.bl)/call.wh
```

```{r}
## what's the average call back rate for each first name
# first, get all first names
names.wh <- unique(d[d$race=="white", "firstname"])
names.bl <- unique(d[d$race=="black", "firstname"])
```

```{r}
# get the mean for a particular name
this.name <- names.bl[1] # first element of the black-sounding names vector
mean(d$call[d$firstname == this.name])
```
\vspace{5mm}

### Using a for loop to analyze the data
We can use a for loop to calculate the average callback rate for each unique name in the data. In general, a for loop is useful in any situation where you want to repeat the same operation over a collection of elements.
\vspace{5mm}
```{r}
## Step 1: Create a vector of elements over which the for loop will iterate
## In this case, we want a vector of unique first names for each subgroup
names.bl <- unique(d$firstname[which(d$race=="black")])
names.wh <- unique(d$firstname[which(d$race=="white")])
```

```{r}
# print the first five elements of each vector to check
names.bl[1:5]
names.wh[c(1,2,3,4,5)]
```
\vspace{5mm}
```{r}
## Step 2: Create an empty 'container object' to save the values of interest
## In this case, we want to save the average callback rate for each name
```

```{r}
# creates vector of 'NA's equal in length to the vectors we created in step 1
call.bl <- rep(NA, length(names.bl))
call.wh <- rep(NA, length(names.wh)) 
```
\vspace{5mm}
```{r}
## Step 3: Specify and run the for loop

## check what a for loop does
# counter variable 'i' will run from 1, 2, ... , 20
for (i in 1:20){
  print(i) # iterates over every value of 'i'
}

## this is some code we will run inside the for loop 
## we will keep it here for debugging
# code for debugging outside of the loop (commented out here)
# i <- 1
# names.bl[i]

## now we want to iterate from 1 to the length of names.bl vector 
## to get every name in the names.bl vector
# counter variable 'i' will run from 1, 2, ... , length(names.bl)
for (i in 1:length(names.bl)) {
  this.name  <- names.bl[i] # what is the i-th name in names.bl
  name.mean  <- mean(d$call[d$firstname==this.name]) # avg callback rate for that name
  call.bl[i] <- name.mean   # save avg callback rate in i-th index of container
}

# can also simplify the above to one line of code
for (i in 1:length(names.wh)) {
  call.wh[i] <- mean(d$call[d$firstname==names.wh[i]])
}

## same task for white-sounding names
for (i in 1:length(names.wh)){
  # take the ith element of 'names.wh' 
  this.name <- names.wh[i]
  # store the mean of callbacks for this.name as the ith element of 
  # call.wh vector
  call.wh[i] <- mean(d$call[d$firstname == this.name])
}
```

\vspace{5mm}
```{r}
# Compare the distributions of average call back rate for black and white names

## Include the first and last lines in this chunk of code if you want
## to save the plot as a PDF file in the current working directory.
## Can also save as bmp(), jpeg(), or png() files (see help manual).

# pdf("plotname.pdf", height = 4, width = 6) [include this line to save plot]
plot(density(call.bl), col="forestgreen",
     xlim=c(0, 0.2), ylim=c(0, 20),
     xlab="Callback Rate", main="Callback Rate by Race")
abline(v=mean(call.bl), lty=2, col="forestgreen")
lines(density(call.wh), col="cornflowerblue")
abline(v=mean(call.wh), lty=2, col="cornflowerblue")
legend("topright", legend=c("'Black' Names", "'White' Names"), 
       col=c("forestgreen", "cornflowerblue"), lwd=1, cex=0.75)
# graphics.off() [include this line to save plot]
```
\vspace{5mm}


```{r}
# get the list of unique first names
firstnames <- unique(d$firstname)

# calculate its number of elements
n <- length(unique(d$firstname))


### Finding call back rates for black sounding names

# step 1: create storage
call.rate <- rep(NA, n) # this is a vector of length n; all elements are NAs

# step 2: loop

## we first do a trial run with the first element ("Allison")

i <- 1 # create an index

# subset the data to all rows whose first names are Allison, calcuate the mean call-back rate
mean(d$call[which(d$firstname == firstnames[i])])

# check whether this is the same as writing Allison directly
mean(d$call[which(d$firstname == "Allison")])

# yes it is! success!

# now we can write a for loop
for (i in 1:n) { # let i run from 1 to n
  # the ith element of the vector call.rate is the average callback rate for the ith first name from the name of unique names we created
  call.rate[i] <- mean(d$call[which(d$firstname == firstnames[i])])
}

# check this vector
call.rate

# we can create a dataframe combining the call rate vector and the first names

cbind.data.frame(firstnames, call.rate)

# or we can give names to elements of this callback rate vector
names(call.rate) <- firstnames
```