---
title: "Introduction to R"
author: "Yiqing Xu (Stanford)"
date: ""
output:
  beamer_presentation:
    incremental: no
  ioslides_presentation:
    transition: 0
    incremental: no
    fig_height: 5.5
    fig_width: 8.2

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Getting Started with ```R```

- Download ```R``` from <https://cran.r-project.org>
- Download ```RStudio``` from <http://www.rstudio.com>
- We will be using ```session1.R``` (code file) and ```nations.RData``` (data file) 

## Big Picture

1. Data are tables, but 
    + Rows and columns
    + There are other types of objects: vectors, matrices
\bigskip  
2. We write R scripts to manipulate data
    + Data need to be loaded and saved
    + R scripts need to be loaded and saved for reproducibility
\bigskip      
3. Packages are collections of generalizable R scripts (functions)
    + Download and install them from CRAN or Github
    + You can write your own functions or packages
\bigskip  
4. You learn to code by coding
    + Breaking things is an essential part of learning
    + Learn with your own pace

## Today's Agenda

- Arithmetic Operations
- Objects
- Vectors
- Functions
- Data Files
- Saving Objects
- Packages
- Programming Tips

## Arithmetic Operations

```R``` can be used as a calculator:

```{r, echo = TRUE}
5 + 3

5 / 3

5 ^ 3
```

## Objects

```R``` is an "object-oriented" programming language. An *object* is any piece of information stored by ```R```. These can be anything, for example:

- A dataset (e.g. a country year dataset)
- A subset of a dataset (e.g. just the democracies in a country year dataset)
- A number (e.g. $2\pi + 1$)
- A phrase (e.g. "Stanford is awesome")
- A function (e.g. a function that takes in $x$ and gives you $x^2 + 8$)

## Objects (cont.)

```R``` can store *objects* with a name of our choice. Use ```<-``` as an assignment operator for objects.

```{r, echo = TRUE}
result <- 5 + 3
result
```

If we assign a new value to the same object name, then we will overwite this object (so be careful when doing so!)

```{r, echo = TRUE}
result <- 5 - 3
result
```

## Objects (cont.)

```R``` can also represent other types of values as objects, such as strings of characters:
```{r, echo = TRUE}
Stanford <- "Stanford is awesome"
Stanford
```

There are many other classes of data besides numeric and character, which we will talk about in class

## Vectors

A *vector* simply represents a collection of information stored in a specific order. We use the function ```c()```, which stands for "concatenate," to enter a data vector (with commas separating elements of the vector):

```{r, echo = TRUE}
world.pop <- c(2525779, 3026003, 3691173, 4449049, 
               5320817, 6127700, 6916183)
world.pop
```

## Vectors (cont.)

To access specific elements of a vector, we use square brackets ```[ ]```. This is called *indexing*:

```{r, echo = TRUE}
world.pop[2]

world.pop[c(2, 4)]
```

## Vectors (cont.)

Since each element of this vector is a numeric value, we can apply arithmetic operations to it:

```{r, echo = TRUE}
world.pop * 1000
```

## Functions

A *function* takes input object(s) and returns an output object. In ```R```, a function generally runs as ```funcname(input)``` where ```funcname``` is the function name and ```input``` is the input object. We often call these inputs *arguments*. Some basic functions useful for summarizing data include:

- ```length()```: length of a vector (number of elements)
- ```min()```: minimum value
- ```max()```: maximum value
- ```range()```: range of data
- ```mean()```: mean
- ```sum()```: sum

## Functions (cont.)

```{r, echo = TRUE}
length(world.pop)

min(world.pop)

max(world.pop)
```

## Functions (cont.)

```{r, echo = TRUE}
range(world.pop)

mean(world.pop)

sum(world.pop)
```

## Data Files

Most of the time, we will load data from an external file. For this class, we will mostly use:

- *CSV*: comma-separated files. These are conceptually similar to Microsoft Excel or Google Spreadsheet.

- *RData*: collection of ```R``` objects including data sets.

## Working Directory

- The *working directory* is where ```R``` will by default load data from and save data to. 

- Use the function ```getwd()``` to display the current working directory.

```{r, echo = TRUE}
getwd()
```

- Use the function ```setwd()``` to change the working directory

```{r, echo = TRUE}
path <- getwd()
setwd(path)
```

- In ```RStudio```, you can also go to Session, Set Working Directory, To Source File Location to set it where your ```R``` file is located

## Reading in Files

- For *CSV* files:

```{r, echo = TRUE}
d <- read.csv("nations.csv")
```

- For *RData* files:

```{r, echo = TRUE}
# d <- load("d.RData")
```

## Data Frames

A *data frame* is a collection of vectors, but we can think of it like a spreadsheet. Useful functions for data frames include:

- ```names()```: return a vector of variable names
- ```nrow()```: return the number of rows
- ```ncol()```: return the number of columns
- ```dim()```: combine ```ncol()``` and ```nrow()``` into a vector
- ```summary()```: produce a summary

## Data Frames (cont.)

```{r, echo = TRUE}
names(d)

nrow(d)

ncol(d)
```

## Data Frames (cont.)

```{r, echo = TRUE}
dim(d)

summary(d)
```

## Data Frames (cont.)

The ```$``` operator is one way to access variables from a data frame:

```{r, echo = TRUE}
d$country
```

Another way of retrieving variables is to use brackets ```[ ]``` with a comma in the form ```[rows, columns]```:

```{r, echo = TRUE}
d[, "gdppc"]

d[1:6, "year"]
```

## Saving Objects

- When you quit ```RStudio```, you will be asked whether you would like to save the workspace. You should answer *no* to this in general: we only want to save what we want!

- To export *CSV*:
```{r, echo = TRUE}
write.csv(d, file = "nations_new.csv")
```

- To export *RData*:
```{r, echo = TRUE}
save(d, file = "nations.RData")
```

## Packages

One of ```R```'s strengths is the existence of a large community of ```R``` users who contribute various functionalities as ```R``` packages. For example, the ```foreign``` package is useful when dealing with files from other statistical software.

```{r, echo = TRUE}
# install.packages("foreign") # install package
library(foreign) # load package
```

## Programming Tips

- First, use the text editor in ```RStudio``` to write your code rather than directly typing it into the R console. That way you can save a record of your program as a ```R``` file.

- Second, we can annotate our ```R``` code so that it is easily understandable to ourselves and others using #

```{r, echo = TRUE}
# File: d.R
# Author: Yiqing Xu
# This code loads the UN population data

d <- read.csv("nations.csv")
```

## Programming Tips (cont.)

Third, we should follow a certain set of coding rules:

- Use informative names for files, variables, and functions
- Use systematic spacing and indentation

Finally, ```R Markdown``` is useful for quickly writing documents using ```R```. I used it to make this presentation, and you will be using this to turn in your problem sets.