---
title: "SRC Summer Camp"
author: "Session 1"
date: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to R

This is an R markdown file summarizing the material from lecture 2. You can also use it as a template for writing your Problem Sets in R markdown. 

First, when writing an R script you can give it a title, date, and name who the author is. 

```{r}
#---
# title: Lecture 2: Introduction to R
# author: Name
# date: Sep 16, 2019
#---
```


## Arithmetic Operations

You can use R as a calculator. You can try summation, division, multiplication, power calculations.

```{r}
5 + 3
5 - 3
5 / 3
5 * 3
5 ^ 3
```



## Objects

R is an "object-oriented" programming language. An *object* is any piece of information stored by R. 
These can be anything, for example:
  
- A dataset (e.g. a country year dataset)
- A subset of a dataset (e.g. just the democracies in a country year dataset)
- A number (e.g. $2\pi + 1$)
- A phrase (e.g. "stanford is awesome")
- A function (e.g. a function that takes in $x$ and gives you $x^2 + 8$)
  
R can store *objects* with a name of our choice. Use <- as an assignment operator for objects.

```{r}
result <- 5 + 3

result
```

If we assign a new value to the same object name, then we will overwite this object (so be careful when doing so!)

```{r}
result <- 5 - 3
result
```

R can also represent other types of values as objects, such as strings of characters:

```{r}
s <- "stanford is awesome"
s
```

There are many other classes of data besides numeric and character, which we will talk about in class.

## Vectors

A *vector* simply represents a collection of information stored in a specific order. 
We use the function c(), which stands for "concatenate," to enter a data vector (with commas separating elements of the vector):

```{r}
world.pop <- c(2525779, 3026003, 3691173, 4449049, 5320817, 6127700, 6916183)
world.pop

a <- c(1, 2, 3, 5)
```

To access specific elements of a vector, we use square brackets [ ]. This is called *indexing*:
  
```{r}
world.pop[2]

world.pop[-1]

world.pop[c(2, 4)]

world.pop[-c(1,3)]
```
  

Since each element of this vector is a numeric value, we can apply arithmetic operations to it:

```{r}
b <- world.pop /1000
```

## Functions

A *function* takes input object(s) and returns an output object. In R, a function generally runs as funcname(input) where funcname is the function name and input is the input object. We often call these inputs *arguments*.

Some basic functions useful for summarizing data include:
  
- length(): length of a vector (number of elements)
- min(): minimum value
- max(): maximum value
- range(): range of data
- mean(): mean
- sum(): sum

```{r}
length(world.pop)

min(world.pop)

max(world.pop)

range(world.pop)[2]

mean(world.pop)

sum(world.pop)
```

## Data Files

Most of the time, we will load data from an external file. For this class, we will mostly use:
  
- *CSV*: comma-separated files. These are conceptually similar to Microsoft Excel or Google Spreadsheet.
- *RData*: collection of R objects including data sets.

## Working Directory

The *working directory* is where R will by default load data from and save data to. You can use the function getwd() to display the current working directory.

```{r}
getwd()
```

You can use the function setwd() to change the working directory

```{r}
# setwd("") # Enter your directory here
```

In RStudio, you can also go to Session, Set Working Directory, To Source File Location to set it where your R file is located.

## Clear memory

You can use the \texttt{rm} function to clear memory or remove objects. 

```{r}
rm(list = ls())
```

## Reading in Files

- For *CSV* files:

```{r}
d <- read.csv("nations.csv")
```

- For *RData* files:
```{r}
rm(list = ls())
load("nations.RData")
```

## Data Frames

A *data frame* is a collection of vectors, but we can think of it like a spreadsheet. 

Useful functions for data frames include:
- names(): return a vector of variable names
- nrow(): return the number of rows
- ncol(): return the number of columns
- dim(): combine ncol() and nrow() into a vector
- summary(): produce a summary

```{r}
names(d)

nrow(d)

ncol(d)

dim(d)

summary(d)
```

The $ operator is one way to access variables from a data frame:
```{r}
table(d$demo)
table(d$year)
```
  
Another way of retrieving variables is to use brackets [ ] with a comma in the form [rows, columns]:
  
```{r}
d[, "country"]

d[1:4, "year"]
```
  
## Saving Objects

When you quit RStudio, you will be asked whether you would like to save the workspace. You should answer *no* to this in general: we only want to save what we want!
  
- To export *CSV*:

```{r}
write.csv(d, file = "nations_new.csv")
```

- To export *RData*:

```{r}
save(d, file = "d.RData")
```

## Packages

One of R's strengths is the existence of a large community of R users who contribute various functionalities as R packages. For example, the foreign package is useful when dealing with files from other statistical software.

```{r}
# install.packages("foreign") # install package
library(foreign) # load package
```

## Programming Tips

First, use the text editor in RStudio to write your code rather than directly typing it into the R console. That way you can save a record of your program as a R file.

Second, we can annotate our R code so that it is easily understandable to ourselves and others using #

```{r}
# File: d.R
# Author: Charlie McClean
# This code loads the UN population data

d <- read.csv("nations.csv")
```

Third, we should follow a certain set of coding rules:

* Use informative names for files, variables, and functions
* Use systematic spacing and indentation

```{r}
# File: d.R
# Author: Charlie McClean
# This code loads the UN population data

d <- read.csv("nations.csv")
```